### 2020-10-26 doc/manuscript/Dow_manuscript.Rmd
Small changes to intro. Add DOYtiming figure

#Figure numbers
schematic=1 #Based on D'Orangeville Figure 1
growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
climwin = 3
pheno_Tsensitivity = 4
model_summary = 5

#SI:
#tables

@@

climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
climwin_dp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for diffuse porous species at SCBI."
```

@@

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018). Therefore, timing of seasonal warming could influence when bud burst occurs, and consequently, lead to variations in total CO2 sequestered by deciduous forests.

Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology traits are dissimilar among trees of different wood types (ring-porous and diffuse-porous). For example, they observed that bud burst was occuring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

Earlier leaf up timings could signal that these trees are shifting their peak growth windows back in response to increasing Spring temperatures, but to date, very little research, has been done on the phenology of stem-growth. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity in temperate forests (Foster, 2014; ). If a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop (source?).

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within a forest dynamics plot in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

>>>>

#Figure numbers
schematic=1 #Based on D'Orangeville Figure 1
growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
climwin = 3
pheno_Tsensitivity = 4
model_summary = 5
DOYtiming = 6
#SI:
#tables

@@

climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
climwin_dp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for diffuse porous species at SCBI."
DOYtiming_caption <- "Figure 6. Day of year where growth milestones were achied at the Smithonian Conservation Biology Institute."
```

@@

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous). For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.

(May remove... not sure where to work in...or if it needs to be in) Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

While the afformentioned broad scale studies have shown a potential relationship between total tree growth and temperature, timing of intraannual stem-growth is extremely important as well. The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; ). Species have co-evolved to utilize the phenological stages of trees at specific times in the year. If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees (). Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).

While not entirely impossible, it is difficult to record intraannual growth traits or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, some traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 15 (5 @ Harvard Forest, 10 @ SCBI) years of biweekly dendrometer band measurements for 19 (15 @ Harvard Forest, 4 @ SCBI) species to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY). Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/DOYtiming.png")
```
